Synthesizing Quotient Lenses

### 2017-11-03 papers/QuotientLensSynth/main.tex

For example, with these rules, the following \bibtex{} citations should be
considered the same by the lens \saz{``the lens'' hasn't been introduced to the
  reader yet},
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }

    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "Conway, J. H.",     }
\end{verbatim}
with the EndNote records having similar rules \saz{similarly, ``rules'' haven't been
  mentioned yet, do you mean similar ``equivalence''} defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences \saz{explicitly mention
  whitespace and permutation?} which nest and interact
with each other in a way that makes them difficult to express with the standard
Boomerang combinators. Moreover, many of the equivalences expressible in
Boomerang require that the programmer first specify the data itself in form of
a regular expression before specifying the equivalence relation on the data.
\saz{Something like this instead?:  Moreover, Boomerang requires that the
  programmer first specify the data itself in the form of a regular expression
  before she can define the equivalence relation on that data.}
This leads to verbose, repetitive code.

To address this issue\saz{these issues? these problems with Boomerang?}, we propose {\em Quotient Regular Expressions} or QREs,
which express a broad class of equivalence relations directly on regular
languages using a compact and convenient notation. For example, with QREs we
can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

@@

\saz{Throughout these examples, refer to the definitions in Figure 1.}  In the
figure, the QRE \cd{wsp_to_space}, which is defined to be \cd{project wsp}$^+$
$\to$ \cd{" "}, matches any non-zero number of whitespace characters, but
considers the matched string of whitespace characters as equivalent with a
single space as the canonical representative.  \saz{We call the type of QRE a
\textit{projection} because it projects an entire regular language down to a single
representative string in the language.}

\saz{
Another form of QRE is the \textit{squash} QRE, an example of which is given by
\cd{valid_name}, which is defined to be:
}

@@

\noindent This QRE specifies that the \cd{comma_name} representation of the
author's name is equivalent to the \cd{space_name} representation.  Here \cd{get
  space_to_comma} is the function that transforms names in the \cd{comma_name}
representation to names in the \cd{space_name} representation.  \saz{In this
  example, \cd{space_to_comma} is itself an automatically \textit{synthesized}
  lens between the \cd{space_name} and \cd{comma_name} QREs, and we need to use
  the ``get'' direction of the lens to specify the squash transformation. (We'll
  say more about synthesis below.)}

@@

\saz{Given the \textit{projection}, \textit{squash}, and \textit{permutation}
  QRE primitives, we can build more complex languages using the usual regular
  expression combinators of alternative choice `\cd{|}', Kleene star `$^*$', and
  sequential composition ``.''.  As shown in the rest of
  Figure~\ref{fig:example}, it is straightforward to define both a regular
  language that accepts \bibtex{} records \textit{and}, simultaneously, an
  equivalence relation that quotients out irrelevant details.}

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) =
   preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between the \cd{bibtex} and \cd{end_note} QRE languages.}
\label{fig:example-lens}
\end{figure}

This is itself a tedious and potentially error-prone task, as the code snippet
in Figure~\ref{fig:example-lens} illustrates.  The lens shown in the Figure
transforms \bibtex{} data to EndNote data  (and vice versa) by matching
the delimiters and separators in a bijective manner. In more detail, the
\cd{preamble} lens replaces the \bibtex{} preamble \verb|@Book| with the EndNote
preamble \verb|%0 Book\n%F|. \saz{This next sentence isn't very well-supported
  by the previous sentence's example, since that doesn't seem too hard.  Maybe
  explain some of the harder parts too?}  The programmer can ill
afford to make a mistake with any one of the small but vital details of the
transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

To address this challenge, we identify a \saz{avoid ``special''} class of quotient
lenses that \saz{instead of ``which''} we call {\em QRE lenses}, and show how to {\em
synthesize} QRE lenses from a pair of QREs and a set of input-output
example pairs. Our key insight is to reduce the problem of synthesizing QRE
lenses to the problem of synthesizing {\em bijective lenses}, which are the
bijections created from the Boomerang lens combinators. This is a problem
which we solved in our prior work ~\cite{popl18}.

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs---\saz{in
  latex use three dashes --- for this (not one -).  Put no spaces around ``---''}every QRE
lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

One potential pitfall with QRE lenses is that the combinators that \saz{instead
  of ``which''}
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ might not be equivalent to
any quotient lens that \saz{instead of ``which''} has the canonical normal form of a bijective lens with
canonizers at the edges. Our main technical contribution in this paper is a
proof that each QRE lenses $q$ defines the same transformation as a QRE lens
$q'$ which is in normal form.

@@

\saz{
As we saw with the definition of \cd{space_to_comma} in
Figure~\ref{fig:example-qre}, the synthesis procedure itself can be used to
creates lenses that are in turn used to define other QREs.  This ability to mix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.}

@@

\saz{One combinator that we haven't yet introduced is sequential composition.}
The $q \semicolon q'$ QRE successively applis
two equivalence relations to a regular language. For example, in the \bibtex{}
to EndNote example, when defining the \cd{valid_name} QRE which maps
\cd{comma_name} into \cd{space_name}, we first defined the equivalence
relations on the \cd{comma_name} and \cd{space_name} representations before
defining the \cd{valid_name} QRE. Using the composition combinator, we could
instead have defined \cd{valid_name} as follows:
\saz{Instead of replaying the \cd{valid_name} example, perhaps use a simpler
  one---even an artificial example might be better.}

@@

The final combinator is  $\normalize{R}{R'}{f}$, which expresses
the equivalence relation that \saz{instead of ``which''} sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

@@

\saz{This is a bit confusing.} \saz{The equivalence relations definable by
  QREs are a subset of those that can be expressed in Boomerang.} 
While QREs express a broad class of equivalence relations on regular languages,
they do not express all the equivalence relations that can be expressed in
Boomerang, nor of course all the equivalence relations which cannot be
expressed using Boomerang canonizers. For instance, Boomerang canonizers can
map a regular language described by $R$ to a regular language described by
$S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $q$
canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$.

@@

More specifically, in order to synthesize a quotient lens $q: c \Leftrightarrow
c'$, that is consistent with the examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$
 it suffices to perform the following procedure:
\begin{enumerate}
  \item
  Compute $K(c)$, $K(c')$,
  \item
  compute $\{({x_1}', {y_1}'), \ldots, ({x_n}', {y_n}')\}$ where ${x_i}' =
  \canonizer(c)(x_i)$ and ${y_i}' = \canonizer(c')(y_i)$,
  \item
    synthesize a bijective lens $\ell : K(c) \Leftrightarrow K(c')$ that is
    consistent with the examples $\{({x_1}', {y_1}'), \ldots, ({x_n}',
    {y_n}')\}$
  \item 
  return $q = \mathit{rquot}(\mathit{lquot}(c, \ell), c') : c \Leftrightarrow
  c'$
\end{enumerate}
As an example, recall that
in the \bibtex{} to EndNote transformation we had the QREs \cd{bibtex} and \cd{endnote} describing a \bibtex{} record and an
Endnote record respectively. We also had the example pair 
\cd{(bib_example , end_example)} where

@@

Since the transformation between the kernel of \cd{bibtex} and the kernel of
\cd{endnote!} that maps \cd{bib_example!} to \cd{end_example} can be
expressed by a bijective lens, it suffices to type the expression

@@

\bcp{No bibliography??}

>>>>

For example, the desired transformation should map the following
\bibtex{} citations to the same EndNote record in the forward direction:
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }

    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "Conway, J. H.",     }
\end{verbatim}
\noindent EndNote records should also have a similar equivalence relation
defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences \saz{explicitly mention
  whitespace and permutation?} which nest and interact
with each other in a way that makes them difficult to express with the standard
Boomerang combinators. Moreover, in many cases, Boomerang requires that the
programmer first specify the data itself in the form of a regular expression
before she can define the equivalence relation on that data. This leads to
verbose, repetitive code.

Instead, we propose {\em Quotient Regular Expressions} or QREs which enable
the programmer to define a regular expression $R$ as well as an equivalence
relation on $R$ in the same expression.For example, with QREs we can specify
\bibtex{} records as shown in Figure~\ref{fig:example-qre}.

@@

In Figure~\ref{fig:example-qre}, the QRE \cd{wsp_to_space}, which is defined to
be

\noindent \cd{project wsp}+ $\to$ \cd{" "}, matches any non-zero number of
whitespace characters, but considers the matched string of whitespace
characters as equivalent with a single space as the canonical representative. 
We call this QRE a \textit{projection} because it projects an
entire regular language down to a single representative string in the
language.

Another form of QRE is the \textit{squash} QRE, an example of which is given by
\cd{valid_name}, which is defined to be:

@@

\noindent This QRE specifies that the \cd{comma_name} representation of the
author's name is equivalent to the \cd{space_name} representation.  Here \cd{get
  space_to_comma} is the function that transforms names in the \cd{comma_name}
representation to names in the \cd{space_name} representation.  In this
  example, \cd{space_to_comma} is itself an automatically \textit{synthesized}
  lens between the \cd{space_name} and \cd{comma_name} QREs, and we need to use
  the ``get'' direction of the lens to specify the squash transformation. (We'll
  say more about synthesis below.)

@@

Given the \textit{projection}, \textit{squash}, and \textit{permutation}
  QRE primitives, we can build more complex languages using the usual regular
  expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
  and Kleene star `$^*$', and also sequential composition ``;''.  As shown in
  the rest of Figure~\ref{fig:example-qre}, it is straightforward to define both a
  regular language that accepts \bibtex{} records \textit{and}, simultaneously,
  an equivalence relation that quotients out irrelevant details.

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) =
   preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between \cd{bibtex} and \cd{end_note} regular
expressions with no equivalences.}
\label{fig:example-lens}
\end{figure}

This is itself a tedious and potentially error-prone task, as the code snippet
in Figure~\ref{fig:example-lens} illustrates.  The lens shown in the Figure
transforms \bibtex{} data to EndNote data  (and vice versa) by matching
the delimiters and separators in a bijective manner. For example, the
\cd{preamble} lens replaces the \bibtex{} preamble \verb|@Book| with the EndNote
preamble \verb|%0 Book\n%F|. When the programmer is writing the lens, she has
to make sure that all of these small deletions and insertions are correct,
which can be a challenging task if the lens is large enough. This makes lens
programming tedious and error-prone.

To address this challenge, we identify a  class of quotient lenses that we call
{\em QRE lenses}, and show how to {\em synthesize} QRE lenses from a pair of
QREs and a set of input-output example pairs. Our key insight is to reduce the
problem of synthesizing QRE lenses to the problem of synthesizing {\em
bijective lenses}, which are the bijections created from the Boomerang lens
combinators. This is a problem which we solved in our prior work ~\cite{popl18}.

QRE lenses lenses are the same as bijective lenses with ``canonizers at the
edges'' as in Figure~\ref{fig:attheedges}, and are typed using QREs---every QRE
lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

@@

One potential pitfall with QRE lenses is that the combinators that we define on
them may not preserve this normal form. For example, the composition $q'; q$ of
QRE lenses $q$ and $q'$ might not be equivalent to any quotient lens that has
the canonical normal form of a bijective lens with canonizers at the edges.
Our main technical contribution in this paper is a proof that each QRE lenses
$q$ defines the same transformation as a QRE lens $q'$ which is in normal form.

@@

As we saw with the definition of \cd{space_to_comma} in
Figure~\ref{fig:example-qre}, the synthesis procedure itself can be used to
creates lenses that are in turn used to define other QREs.  This ability to mix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

One combinator that we haven't yet introduced is sequential composition.
The $q \semicolon q'$ QRE successively applis
two equivalence relations to a regular language. For example, in the \bibtex{}
to EndNote example, when defining the \cd{valid_name} QRE which maps
\cd{comma_name} into \cd{space_name}, we first defined the equivalence
relations on the \cd{comma_name} and \cd{space_name} representations before
defining the \cd{valid_name} QRE. Using the composition combinator, we could
instead have defined \cd{valid_name} as follows:
\saz{Instead of replaying the \cd{valid_name} example, perhaps use a simpler
  one---even an artificial example might be better.}

@@

The final combinator is  $\normalize{R}{R'}{f}$, which expresses
the equivalence relation that sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

@@

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can map a
regular language described by $R$ to a regular language described by $S$, with
$\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $q$ canonizes
a regular language described by $R$ to regular subset $\mathcal{L}(R')
\subseteq \mathcal{L}(R)$ described by $R'$.

@@

More specifically, in order to synthesize a quotient lens $q: c \Leftrightarrow
c'$, that is consistent with the examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$
 it suffices to perform the following procedure:
\begin{enumerate}
  \item
  Compute $K(c)$, $K(c')$,
  \item
  compute $\{({x_1}', {y_1}'), \ldots, ({x_n}', {y_n}')\}$ where ${x_i}' =
  \canonizer(c)(x_i)$ and ${y_i}' = \canonizer(c')(y_i)$,
  \item
    synthesize a bijective lens $\ell : K(c) \Leftrightarrow K(c')$ that is
    consistent with the examples $\{({x_1}', {y_1}'), \ldots, ({x_n}',
    {y_n}')\}$
  \item 
  return $q = \mathit{rquot}(\mathit{lquot}(c, \ell), c') : c \Leftrightarrow
  c'$
\end{enumerate}
For example, recall that in the \bibtex{} to EndNote transformation we had the
QREs \cd{bibtex} and \cd{endnote} describing a \bibtex{} record and an Endnote
record respectively. We also had the example pair \cd{(bib_example ,
end_example)} where

@@

Since the transformation between the kernel of \cd{bibtex} and the kernel of
\cd{endnote} that maps \cd{bib_example} to \cd{end_example} can be
expressed by a bijective lens, it suffices to type the expression

@@

\bibliography{local}
